What Can Stack Traces Tell us about the Use of Exceptions in Java?

### 2014-07-02 android-stacks.tex
improving process description

\noindent\emph{Checked Approach}. In languages such as CLU \footnote{In 
CLU if a handler is found for a given exception, it is converted in a predened 
default exception, called failure which is the only one which can be implicitly 
propagated in this model.}, Modula-3, Guide, and Extended Ada all exceptions 
are checked, which means that the compiler
statically checks if appropriate handlers are provided within the system.
Accordingly, the programmer must explicitly specify every exception that can be
signaled to support the compile time checking. This approach is known to be
costly to maintain (Dooren and Steegmans, 2005) since exception updates are
cascading; if a new exception type is added to a method signature then all other
methods directly using this method must either change their signatures to
include the new exception type or handle the new exception.

\noindent\emph{Unchecked Approach}. In languages such as Lore, C++, and Arche
all exceptions are unchecked but the developer can optionally list exception
types on method signatures. Although in some languages such exceptions can 
be documented on the methods signature, the developer is not warned by the compiler
if an unchecked exception is not handled inside the application. According to ~\cite{Robil00} 
such an approach may hinder robustness because the client of a method cannot easily know
which unchecked exceptions may be thrown (unless the code of the method and the
methods called from it is inspected) which can be a very time consuming or
infeasible task.

\noindent\emph{Hybrid Approach}. In Java a method can throw either checked or
unchecked exceptions. Java is the only language that allows this hybrid approach,
on which the developer can define exceptions of both types of exceptions to be thrown and
handled in a single system.

@@

%\subsubsection{Exception Chaining} Exception Chaining 
%This stack trace was extracted from an issue of an Android application in
%Github.

Exception chaining is a common way of propagating exceptions in Java programs, 
which allows one exception to be wrapped in another exception and re-thrown. 
Figure~\ref{fig:wrapping} presents a stack trace associated which illustrates an
exception chaining. The bottom part of the stack trace is the \emph{root cause}, which indicates the
first reason for the error to be thrown (in this case, the computer run out of
memory). The top part of the stack trace indicates the location of the exception
manifestation (which will call \emph{top level exception} along this paper). The
execution flow  between the root cause and the exception manifestation may
include other intermediate exception wrappings. In all levels, the exception
\emph{signaler}, is the method that threw the exception, represented on the
stack trace as the first method call below the exception declaration.

@@

For RQ1 and RQ2, we explore the domain quantitatively, and use theirs answers as
a basis to investigate the third research question. RQ3 aims at examining whether
exception (mis)-uses could be revealed through stack trace analysis. To
support this investigation, source code and bytecode analysis was used to
leverage the  understanding of stack traces. 
%For RQ3 we investigate
%quantitatively and highlight interesting cases by exploring cases qualitatively.
Figure~\ref{fig:overviewfig} presents an overview of the mixed-methods approach
conducted to answer these questions.

@@

We begun our investigation using the dataset provided by the GHTorrent
project~\cite{Gousi13}, an off-line mirror of the data offered through the
Github API. The Github API data come in two flavours; a streaming data flow lists
events, such as reporting an issue, happening on repositories in real time,
while a static view contains the current state of entities. The data is stored
in unprocessed format, in a MongoDB database, while metadata is extracted and
stored in a relational database. The GHTorrent dataset covers a broad range of
development activities on Github, including issues. The project has been
collecting data since February 2012. Up to Dec 2013, when we queried it to
retrieve our bootstrapping sample, it included 356.057 issues from 16.836
Java projects (step 1).\footnote{A Java project is specified in GitHub with a
specific label.}

After obtaining the bootstrap issue sample from GHTorrent, each issue was
processed by ExceptionMiner tool developed in this work which extracted every
exception defined on issues or on issue comments (step 2). Since this initial
set of Java project is too diverse (it includes libraries, web-based or
stand-alone applications), in the second phase of this study we restrict the
analysis domain to Android projects (step 3). This decision was driven by the
fact that most open source Android applications are known for dealing
with several sources of exceptions due to multi-threaded execution and
interaction with input/output resources; moreover they share a common
well-defined platform and a limited set of libraries.

@@

To better investigate the extent of this problem we classified each stack trace according to the type of the
root exception. Table~\ref{tab:typeroottab} presents the types of root causes of all stack traces reported on 
the Android repositories. As we can see from Table ~ref{typeroottab},  most of the exceptions signaled by third-party code
 (e.g., Android and libs) were runtime exceptions. After filtering all the exceptions implicitly signaled by 
JVM (due to programming mistakes) and inspecting the signaler methods of such exceptions (i.e.,  direct an inderect 
instances of java.lang.RuntimeException). We could observe that only 1 method out of 118 inspected signalers
 (i.e., 0,8\%) documented the explicitly thrown runtime exception in a Javadoc comment; and none of these methods
included the runtime exception  (reported on the stack) as part of the exception interface of the method (i.e., using 
throws clause on method signature). This result is aligned with the results of other  study conducted by from 
Sacramento et al ~\cite{sacramento2006unchecked} which observed that the
runtime exceptions in .Net programs are most often not documented.

@@

The exception (mis-)use patterns that emerged  in this study sheds light on 
problems that may be reducing the robustness of current Java programs in the face of
exceptional conditions. \noindent\emph{The undocumented runtime exceptions} thrown by library code -
documenting runtime exceptions is a tedious and error prone task, to help developers
mitigate this problem tools should be developed to automate the documentation of runtime exceptions
scraping from library code, few solutions in this directions have been proposed so far ~\cite{van2005combining}. 
 \noindent\emph{The odd wrappings}  found in this study can lead to what we call
an \emph{exception handling confusion problem} which can lead the program to
an unpredictable state in the presence of exceptions. To illustrate this problem, we can use
 one of the examples found in this study: when the developer is 
confronted with a checked exception, the designer of the API is telling him 
to handle the exceptional condition, however we could observe that in some cases the 
checked exception wrapped an OutOfMemoryError, which represent resource deficiency detected 
by the JVM which make impossible the program to recover. Further investigation is needed on finding 
ways to help developers in deadling with such problem, either preventing odd wrappings or enabling 
the developer to better dealing with them. Last but not least, the high number of null pointer exceptions 
lead us to think on other pertinent research direction: whether language designs which avoid null pointers, 
such as Monads ~cite{Walde95} (i.e., used in functional languages for values that may not be available 
or computations that may fail) could improve the robustness of Java programs.

>>>>

In languages such as CLU, Modula-3, Guide, and Extended Ada the exceptions are 
checked, which means that the 
compiler statically checks if appropriate handlers are provided within the system. 
Accordingly, the programmer must explicitly specify every exception that can be
signaled to support the compile time checking \footnote{In CLU,  if a handler is 
found for a given exception, it is converted in a predened default exception, 
called failure which is the only one which can be implicitly 
propagated in this model.}. This approach is known to be
costly to maintain (Dooren and Steegmans, 2005) since exception updates are
cascading; if a new exception type is added to a method signature then all other
methods directly using this method must either change their signatures to
include the new exception type or handle the new exception.

In languages such as Lore, C++, and Arche all exceptions are unchecked but the developer can optionally list exception types on method signatures. Although exceptions can 
be documented on methods signatures, the developer is not warned by the compiler
if an unchecked exception is not handled inside the application. According to ~\cite{Robil00} 
such an approach may hinder robustness because the client of a method cannot easily know
which unchecked exceptions may be thrown (unless the code of the method and the
methods called from it is inspected) which can be a very time consuming or
infeasible task.

In Java a method can throw either checked or unchecked exceptions. Java is the only language that allows this hybrid approach, on which the developer can define exceptions of both kinds to be thrown and
handled in a single system. Next section details Java exception handling model.

@@

%\subsubsection{Exception Chaining} Exception Wrapping 
%This stack trace was extracted from an issue of an Android application in
%Github.

The Exception wrapping (also known as exception chaining) is a common way of 
propagating exceptions in Java programs, 
which allows one exception to be wrapped in another exception and re-thrown. 
Figure~\ref{fig:wrapping} presents a stack trace associated which illustrates an
exception wrapping. The bottom part of the stack trace is the \emph{root cause}, which indicates the
first reason for the error to be thrown (in this case, the computer run out of
memory). The top part of the stack trace indicates the location of the exception
manifestation (which will call \emph{top level exception} along this paper). The
execution flow  between the root cause and the exception manifestation may
include other intermediate exception wrappings. In all levels, the exception
\emph{signaler}, is the method that threw the exception, represented on the
stack trace as the first method call below the exception declaration.

@@

For RQ1 and RQ2, we explore the domain quantitatively, and use their answers as
a basis to investigate the third research question. RQ3 aims at examining whether
exception (mis)-uses could be revealed through stack trace analysis. To
support this investigation, source code and bytecode analysis was used to
leverage the  understanding of stack traces. 
%For RQ3 we investigate
%quantitatively and highlight interesting cases by exploring cases qualitatively.
Figure~\ref{fig:overviewfig} presents an overview of the mixed-methods approach
conducted to answer these questions.

@@

We begun our investigation using the dataset provided by the GHTorrent
project~\cite{Gousi13}, an off-line mirror of the data offered through the
Github API. The Github API data come in two flavours; a streaming data flow lists
events, such as reporting an issue, happening on repositories in real time,
while a static view contains the current state of entities. The data is stored
in unprocessed format, in a MongoDB database, while metadata is extracted and
stored in a relational database. The GHTorrent dataset covers a broad range of
development activities on Github, including issues. The project has been
collecting data since February 2012. Up to Dec 2013, when we queried it to
retrieve our sample, it included  16.836 Java projects (step 1) which overall contained
356.057 issues.\footnote{A Java project is specified in GitHub with a
specific label.}

After obtaining this sample of projects and associated issues, 
the ExceptionMiner tool, developed in this work, performed a per-project 
mining task as follows: for each project, the content of each issue and issue comments 
were analized and all exceptions and exception stack traces were extracted (step 2). 
Since this initial set of Java project is too diverse (it includes libraries, web-based or
stand-alone applications), in the second phase of this study we restrict the
analysis domain to Android projects (step 3). This decision was driven by the
fact that most open source Android applications are known for dealing
with several sources of exceptions due to multi-threaded execution and
interaction with input/output resources; moreover they share a common
well-defined platform and a limited set of libraries.

@@

To better investigate the extent of this problem we classified each stack trace according to the type of the
root exception. Table~\ref{tab:typeroottab} presents the types of root causes of all stack traces reported on 
the Android repositories. As we can see from Table ~\ref{tab:typeroottab},  most of the exceptions signaled by third-party code
 (e.g., Android and libs) were runtime exceptions. After filtering all the exceptions implicitly signaled by 
JVM (due to programming mistakes) and inspecting the signaler methods of such exceptions (i.e.,  direct an inderect 
instances of java.lang.RuntimeException). We could observe that only 1 method out of 118 inspected signalers
 (i.e., 0,8\%) documented the explicitly thrown runtime exception in a Javadoc comment; and none of these methods
included the runtime exception  (reported on the stack) as part of the exception interface of the method (i.e., using 
throws clause on method signature). This result is aligned with the results of other  study conducted by from 
Sacramento et al ~\cite{sacramento2006unchecked} which observed that the
runtime exceptions in .Net programs are most often not documented.

@@

The exception (mis-)use patterns that emerged  in this study sheds light on 
problems that may be reducing the robustness of current Java programs in the face of
exceptional conditions. \noindent\emph{The undocumented runtime exceptions} thrown by library code -
documenting runtime exceptions is a tedious and error prone task, to help developers
mitigate this problem tools should be developed to automate the documentation of runtime exceptions
scraping from library code, few solutions in this directions have been proposed so far ~\cite{van2005combining}. 
 \noindent\emph{The odd wrappings}  found in this study can lead to what we call
an \emph{exception handling confusion problem} which can lead the program to
an unpredictable state in the presence of exceptions. To illustrate this problem, we can use
 one of the examples found in this study: when the developer is 
confronted with a checked exception, the designer of the API is telling him 
to handle the exceptional condition, however we could observe that in some cases the 
checked exception wrapped an OutOfMemoryError, which represent resource deficiency detected 
by the JVM which make impossible the program to recover. Further investigation is needed on finding 
ways to help developers in deadling with such problem, either preventing odd wrappings or enabling 
the developer to better dealing with them. Last but not least, the high number of null pointer exceptions 
lead us to think on other pertinent research direction: whether language designs which avoid null pointers, 
such as Monads ~\cite{Walde95} (i.e., used in functional languages for values that may not be available 
or computations that may fail) could improve the robustness of Java programs.
